Bird beak's elaboration and exploration

### 2021-09-01 Analysis/03-elaboration_exploration_analyses.Rmd
pipeline rework

```{r, message = FALSE, warning = FALSE, fig.height = 5, fig.width = 5}
set.seed(42)
## Space plot
cor_matrix <- matrix(cbind(1, 0.8, 0.8, 1), nrow = 2)
space <- space.maker(50, 2, rnorm, cor.matrix = cor_matrix)
lim <- c(floor(range(space)[1]), ceiling(range(space)[2]))
plot(space, pch = 19, xlab = "Trait 1", ylab = "Trait 2", xlim = lim, ylim = lim)

## Plotting the ellipse
lines(ellipse::ellipse(cor_matrix), col = "blue", lwd = 3)
## Plotting the major axis
lines(get.axes(list(list(list(VCV = cor_matrix, Sol = c(0,0)))))[[1]][[1]], lwd = 3, col = "orange")
```

@@

```{r, fig.height = 6, fig.width = 6}
sauron.plot(covar_matrices,
            col = colour_vector,
            main = "Average ellipses and\nmajor axes",
            major.axes = mean,
            ellipses = mean,
            legend = TRUE)
```

@@

Figure: in a space with five elements: A, B, C, D, E (in grey); where D and E represent the major axes (e.g. the major phylogenetic effect axes from the MCMCglmm), we can rotate and rescale each element so that D and E become the unit vector of length 1 (the black letters D and E) and get the exploration and elaboration scores for either the elements (e.g. element A's projection in blue and its rejection in orange) or from any other axes (e.g. the major phylogenetic effect axes from the MCMCglmm for a specific clade).

@@

```{r}
## Get all the angles and stuff
group_results <- analyses.group(major_axes, base = "animal")
names(group_results) <- levels(demo_data$clade)
```

@@

```{r}
## Plot the blob wise metrics
plot.analyses.group(group_results, col = colour_vector)
```

@@

```{r}
## Global analyses
tip_results_global <- analyses.tip(data = demo_data[, c(1:4)],
                                   axes = major_axes$animal)
```

@@

```{r}
## Plot results
plot.analyses.tip(tip_results_global,
                  col = colour_vector,
                  group = split(rownames(demo_data), f = demo_data$clade),
                  main = "Tips vs. whole phylogeny")
```

@@

```{r}
## Results within groups
tip_results_within <- analyses.tip(
                    data = demo_data[, c(1:4)],
                    axes = major_axes[-length(major_axes)],
                    group = split(rownames(demo_data), f = demo_data$clade))
```

```{r}
## Plot results
plot.analyses.tip(tip_results_within,
                  col = colour_vector,
                  group = split(rownames(demo_data), f = demo_data$clade),
                  main = "Tips vs. clade")
```

>>>>

```{r, message = FALSE, warning = FALSE, fig.height = 5, fig.width = 5}
get.one.axis <- function(data, axis = 1, level = 0.95, dimensions) {

    # The magic: https://stackoverflow.com/questions/40300217/obtain-vertices-of-the-ellipse-on-an-ellipse-covariance-plot-created-by-care/40316331#40316331

    ## Select the right dimensions
    data$VCV <- data$VCV[dimensions, dimensions, drop = FALSE]

    ## Get the data dimensionality
    dims <- dim(data$VCV)[1]

    ## Create the unit hypersphere (a hypersphere of radius 1) for the scaling
    unit_hypersphere1 <- unit_hypersphere2 <- matrix(0, ncol = dims, nrow = dims)
    ## The "front" (e.g. "top", "right") units
    diag(unit_hypersphere1) <- 1
    ## The "back" (e.g. "bottom", "left") units
    diag(unit_hypersphere2) <- -1
    unit_hypersphere <- rbind(unit_hypersphere1, unit_hypersphere2)
    ## Scale the hypersphere (where level is the confidence interval)
    unit_hypersphere <- unit_hypersphere * sqrt(qchisq(level, 2))

    ## Do the eigen decomposition (symmetric - faster)
    eigen_decomp <- eigen(data$VCV, symmetric = TRUE)

    ## Re-scaling the unit hypersphere
    scaled_edges <- unit_hypersphere * rep(sqrt(eigen_decomp$values), each = dims*2)
    ## Rotating the edges coordinates
    edges <- tcrossprod(scaled_edges, eigen_decomp$vectors)

    ## Move the matrix around
    edges <- edges + rep(data$Sol[dimensions, drop = FALSE], each = dims*2)

    ## Get the edges coordinates
    return(edges[c(axis, axis+dims), , drop = FALSE])
}

set.seed(42)
## Space plot
cor_matrix <- matrix(cbind(1, 0.8, 0.8, 1), nrow = 2)
space <- space.maker(50, 2, rnorm, cor.matrix = cor_matrix)
lim <- c(floor(range(space)[1]), ceiling(range(space)[2]))
plot(space, pch = 19, xlab = "Trait 1", ylab = "Trait 2", xlim = lim, ylim = lim)

## Plotting the ellipse
lines(ellipse::ellipse(cor_matrix), col = "blue", lwd = 3)
## Plotting the major axis
lines(get.one.axis(list(VCV = cor_matrix, Sol = c(0,0))), lwd = 3, col = "orange")
```

@@

```{r, fig.height = 6, fig.width = 6}
sauron.plot(covar_matrices,
            col = colour_vector,
            main = "Average ellipses and\nmajor axes",
            major.axes = mean,
            ellipses = mean,
            legend = TRUE,
            cex = 0.5)
```

@@

Figure: consider a space with five elements named "a" to "e" (in grey) and one focal axes in this space defined as the vector from point1: "d" to point2: "e" (the thin solid grey line linking the grey "d" and "e"). We can then define all points from "d" (point1) as vectors (the dashed grey lines). We can then rotate the whole space so that the vector "d" to "e" becomes the reference coordinate vector (0,1) (the thick grey line) and rotate and translate all the other vectors accordingly (the solid grey lines linking the coloured letters "a" to "c"). We can then project each of these vectors on the reference vector and measure their distance *on* the reference vector (the projection; the coloured solid lines) and their distance *from* the reference vector (the rejection; the coloured dotted lines). We can then also measure the angles between each vector and their respective projections (the angles between a grey solid line to a coloured letter and the corresponding coloured vector). Note that if this angle is equal to 90 degrees, the projection is equal to 0 (point "c") or if that angle is > 90 degrees, the projection is negative (point "b").

@@

```{r, fig.height = 6, fig.width = 18}
## Projecting each group's major axes on the phylogenetic major axes
group_elaboration <- dispRity.covar(covar_matrices,
               metric = projections.covar,
               between.groups = list(c(1,4), c(2,4), c(3,4)),
               measure = "position")
group_exploration <- dispRity.covar(covar_matrices,
               metric = projections.covar,
               between.groups = list(c(1,4), c(2,4), c(3,4)),
               measure = "distance")
group_angle <- dispRity.covar(covar_matrices,
               metric = projections.covar,
               between.groups = list(c(1,4), c(2,4), c(3,4)),
               measure = "degree")
par(mfrow = c(1,3))
plot(group_elaboration, ylab = "elaboration", col = colour_vector[-4])
plot(group_exploration, ylab = "exploration", col = colour_vector[-4])
plot(group_angle, ylab = "angle", col = colour_vector[-4])
```

@@

@@

```{r, eval = FALSE}
## Get the mean axes for each level
major_axes <- axis.covar(covar_matrices, sample = mean)

## Projecting each group's major axes on the phylogenetic major axes
phy_group_elaboration <- dispRity(covar_matrices,
               dimensions = covar_matrices$call$dimensions,
               metric  = dispRity::projections,
               measure = "position",
               point1  = major_axes$phylogeny[[1]][1,],
               point2  = major_axes$phylogeny[[1]][2,])
phy_group_exploration <- dispRity(covar_matrices,
               dimensions = covar_matrices$call$dimensions,
               metric  = dispRity::projections,
               measure = "distance",
               point1  = major_axes$phylogeny[[1]][1,],
               point2  = major_axes$phylogeny[[1]][2,])
phy_group_angle <- dispRity(covar_matrices,
               dimensions = covar_matrices$call$dimensions,
               metric  = dispRity::projections,
               measure = "degree",
               point1  = major_axes$phylogeny[[1]][1,],
               point2  = major_axes$phylogeny[[1]][2,])

## TODO: add option in dispRity plot to only select some groups
par(mfrow = c(1,3))
plot(phy_group_elaboration, col = colour_vector)
plot(phy_group_exploration, ylab = "exploration", col = colour_vector)
plot(phy_group_angle, ylab = "angle", col = colour_vector)

## Global analyses
tip_results_global <- analyses.tip(data = demo_data[, c(1:4)],
                                   axes = major_axes$animal)
```

@@

```{r, eval = FALSE}
## Plot results
plot.analyses.tip(tip_results_global,
                  col = colour_vector,
                  group = split(rownames(demo_data), f = demo_data$clade),
                  main = "Tips vs. whole phylogeny")
```

@@

```{r, eval = FALSE}
## Results within groups
tip_results_within <- analyses.tip(
                    data = demo_data[, c(1:4)],
                    axes = major_axes[-length(major_axes)],
                    group = split(rownames(demo_data), f = demo_data$clade))
```

```{r, eval = FALSE}
## Plot results
plot.analyses.tip(tip_results_within,
                  col = colour_vector,
                  group = split(rownames(demo_data), f = demo_data$clade),
                  main = "Tips vs. clade")
```
